# (PART) Modules {-}

# Module 1: Data Cleaning and Exploration Review

## Lecture

<iframe width="640" height="360" src="https://www.youtube.com/embed/FGTl27bddFI?si=HEozmRz4dyfTycqZ" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture; web-share" referrerpolicy="strict-origin-when-cross-origin" allowfullscreen></iframe>

<br>

<iframe src="https://drive.google.com/file/d/13NcJR_WhQz1vYBAjwqbNY8aFlAJA11YK/preview" width="640" height="480" allow="autoplay"></iframe>  

## Lab

### Load Libraries ----

To install a package:

```{r, message=FALSE, warning=FALSE}
install.packages("tidyverse")
```

To load a library:

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(rfishbase)
library(factoextra)
```

### Read in Data ----

Get your current working directory.

```{r, results='hide', message=FALSE, warning=FALSE}
getwd()
```

Download csv file from this link: https://doi.org/10.5061/dryad.f6t39kj
Place it into your current working directory.

Read in data.

```{r, message=FALSE, warning=FALSE}
dfReptiles <- read.csv('Appendix S1 - Lizard data version 1.0.csv')
```

### Looking at your data. ----

```{r, message=FALSE, warning=FALSE}
# What class is it?
class(dfReptiles)

# How many rows and columns?
dim(dfReptiles)

# Look at the column names.
colnames(dfReptiles)
names(dfReptiles)[1:10]

# If you want to view your dataset.
head(dfReptiles, 10)

# We have a lot of columns. Let's subset to remove columns we don't need for now.
dfTraits <- dfReptiles %>% ## tidyverse pipe 
  # Select function allows us to choose the columns we want
  select(c(Binomial, Genus, Family,
           main.biogeographic.Realm, Latitude.centroid..from.Roll.et.al..2017.,
           insular.endemic, maximum.SVL, hatchling.neonate.SVL, Leg.development,
           Activity.time, substrate, diet, foraging.mode, reproductive.mode,
           smallest.clutch, largest.clutch, youngest.age.at.first.breeding..months.,
           IUCN.redlist.assessment, IUCN.population.trend, Extant.Extinct
           ))

names(dfTraits)
# Let's clean up these column names!
names(dfTraits) <- tolower(names(dfTraits))

# Replace all "." with "_" (personal preference)
names(dfTraits) <- gsub("\\.", "_", names(dfTraits))
names(dfTraits)
```

```{r, results='hide', message=FALSE, warning=FALSE}
# Make some of the names shorter.
names(dfTraits)[5] <- "latitude"
names(dfTraits)[17] <- "age_first_breeding"
names(dfTraits)[17]

# Rename species column.
names(dfTraits)[1] <- "species"

# In order to properly count the number of missing values, replace blanks with NAs
# I always do this just in case.
dfTraits[dfTraits == " "] <- NA

# Make sure there are no species name duplications. 
# Note that using the sum() function on logical vector will count the number of TRUE values.
duplicated(dfTraits$species)
sum(duplicated(dfTraits$species))
```

```{r, message=FALSE, warning=FALSE}
# Are there any species that don't have ANY data?
missRows <- apply(dfTraits[, -(1:3)], MARGIN = 1, function(x) all(is.na(x)))
# Let's break that down! First, we ed if "x" row had NAs.
is.na(dfTraits[1, ]) ## logical vector
# Then we ed if they were ALL NAs (it is ing if all of the elements are NA)
dfTraits[1, ]
all()
# Note you could use the "any" function to  if ANY of the elements are NA
any() ## There are a couple NAs
```

```{r, results='hide', message=FALSE, warning=FALSE}
# Then we "apply" that function to all of the rows (MARGIN = 1) in the dataframe which returns:
missRows
```

```{r, message=FALSE, warning=FALSE}
sum(missRows)

# If you wanted to remove species without any trait data from the dataframe:
dfTraits <- dfTraits[!missRows, ]
```

### Working with different data types. ----

Let's try to understand the type of data we are working with.

```{r, message=FALSE, warning=FALSE}
names(dfTraits)

# First, let's ID which columns are taxonomic information so we don't include them in our summary stats.
taxCols <- c("species", "genus", "family")

# The rest are traits
traits <- setdiff(names(dfTraits), taxCols)
traits

#  the classes of the traits.
sapply(dfTraits[traits], class)

# Let's ID which traits are numerical and which are categorical.
index <- sapply(dfTraits[traits], is.numeric)
index

# Wait! Something doesn't seem right..! 
class(dfTraits$maximum_svl)
head(dfTraits)

# The column has strings mixed with numbers, which returns a character vector.
# We should replace the strings with NAs.
# Here, we are using regex to match any letter and then replacing the matches with NAs.
dfTraits$maximum_svl <- gsub(pattern = "[a-zA-Z]", replacement = NA, x = dfTraits$maximum_svl)
dfTraits$hatchling_neonate_svl <- gsub("[a-zA-Z]", NA, dfTraits$hatchling_neonate_svl)

# Change both traits to numeric.
# Note here I am using lapply to apply the function for the columns of dfTraits.
# Note: lapply returns a list, sapply returns a vector. "map" would be the tidyverse equivalent of apply, lapply, sapply, etc. functions.
dfTraits[, c("maximum_svl", "hatchling_neonate_svl")] <- lapply(dfTraits[, c("maximum_svl", "hatchling_neonate_svl")], as.numeric)

class(dfTraits$maximum_svl)

# Let's try IDing our numerical traits again.
index <- sapply(dfTraits[traits], is.numeric)

# Subset the column names using indexing.
contTraits <- traits[index]
contTraits ## is this right?

# Get the categorical traits.
catTraits <- setdiff(traits, contTraits)

# Convert character to factor because this is helpful for summary stats and plotting.
# It is also the data class that regression models require for categorical variables.
dfTraits[catTraits] <- lapply(dfTraits[catTraits], as.factor)
# If you wanted to make a particular category within a variable the reference.
# Important for statistical analyses with categorical variables.
table(dfTraits$insular_endemic)
dfTraits$insular_endemic <- relevel(dfTraits$insular_endemic, "no")

# One last .
sapply(dfTraits, class)
```

### Summary stats. ----

Some base R summary statistics to get a quick look at your data.

```{r, message=FALSE, warning=FALSE}
summary(dfTraits)
```

Time for some tidyverse magic!

Use of the tidyverse pipe avoids us having to create several interim variables while also improving readability.

```{r, message=FALSE, warning=FALSE}
# Say we only wanted to keep extant species:
dfTraits <- dfTraits %>%
  # Filter function allows us to apply a condition to our data.
  filter(extant_extinct == "extant") %>%
  # Remove column using minus sign as we no longer need it
  select(-extant_extinct)

# Remove trait from catTraits.
catTraits
catTraits <- catTraits[-11]
  
# Other summary info.
# How many families do we have?
unique(dfTraits$family)
length(unique(dfTraits$family))

# Top 10 Families with most species.
# You can sort by increasing or decreasing number of species.
head(sort(table(dfTraits$family), decreasing = T), n = 10)

# Doing the same thing, but with tidyverse syntax
dfTraits %>%
  dplyr::count(family, sort = T) %>% ## to make sure the count function isn't masked
  head(n = 10)

# Sample size (number of complete observations for this trait)
head(na.omit(dfTraits$maximum_svl)) ## na.omit removes NAs from the vector
length(na.omit(dfTraits$maximum_svl))

# Mean of the data
mean(dfTraits$maximum_svl, na.rm = T) ## has option for removing number of NAs in the function

# Range of the data
range(dfTraits$maximum_svl, na.rm = T)

# Proportion of NAs
sum(is.na(dfTraits$maximum_svl))
sum(is.na(dfTraits$maximum_svl)) / length(dfTraits$maximum_svl)

# To speed things up, (l)apply these to all of the numerical traits using an anonymous function.
# An anonymous function is a function without a name that you really only need temporarily
# e.g., within the confines of this lapply call.
l_contInfo <- lapply(dfTraits[contTraits], function(x){
  
  # Number of complete observations
  n <- length(na.omit(x))
  
  # Mean
  avg <- mean(x, na.rm = T)
  
  # Number of NAs
  numNAs <- sum(is.na(x))
  
  # Proportion of NAs
  propNAs <- sum(is.na(x)) / length(x)
  
  # Return in dataframe format
  return(data.frame(n, avg, numNAs, propNAs))
  
})

# View the first element of the list.
head(l_contInfo[[1]])

# Bind list of dataframes together by using the do.call() function.
# This lets you rbind() the entire list of dataframes.
dfContInfo <- do.call(rbind, l_contInfo)

head(dfContInfo)

# Do the same thing for the categorical data.
lapply(dfTraits[catTraits], table)

l_catInfo <- lapply(dfTraits[catTraits], function(x){
  
  n <- length(na.omit(x))
  # number of unique categories instead of mean for example
  cats <- length(unique(x))
  numNAs <- sum(is.na(x))
  propNAs <- sum(is.na(x)) / length(x)
  
  return(data.frame(n, cats, numNAs, propNAs))
  
})

# Bind together list.
dfCatInfo <- do.call(rbind, l_catInfo)

head(dfCatInfo)

# Tidyverse has handy functions for getting summary data by group.
# For example, if we wanted to get summary information grouped by family:
summary_stats1 <- dfTraits %>%
  # Group by family
  group_by(family) %>%
  # Get the mean max SVL for each group and put it into a new column called avg_length
  summarize(avg_length = mean(maximum_svl, na.rm = T)) %>%
  # Arrange in descending order
  arrange(desc(avg_length)) %>%
  # Print to console
  print()

# Let's add some info on other traits
summary_stats2 <- dfTraits %>%
  group_by(family) %>%
  summarize(
    avg_length = mean(maximum_svl, na.rm = T),
    # Average largest clutch
    avg_lc = mean(largest_clutch, na.rm = T),
    # Most common diet in each family
    top_diet = names(sort(table(diet), decreasing = T)[1]),
  ) %>%
  print()
```

### Exploratory plots. ----

Let's perform some data visualization to identify patterns and variable associations in our dataset.

```{r, message=FALSE, warning=FALSE}
# Base R histograms for numerical traits
hist(dfTraits$maximum_svl) ## there are some VERY long species!
hist(dfTraits$latitude)

# ggplot to beautify the data
# ggplot is a very powerful tool for visualizing data but you need to get used to the syntax
ggplot(dfTraits) + ## note use of "+" over "%>%
  # Plot a histogram and make it blue.
  # geom_* indicates what type of plot you want. aes = aesthetic mapping
  geom_histogram(mapping = aes(x = latitude), 
                 fill = "skyblue", colour = "black") +
  # Add labels
  labs(title = "Reptile Latitude", 
       x = "Latitude (°)", y = "Count") +
  # Change the plot theme (here, making the background white)
  theme_minimal(base_size = 12) + 
  # Change the theme and make some font adjustments
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

# barplot for categorical traits
plot(dfTraits$diet)

# ggplot version
# Get rid of those pesky NAs for plotting
ggplot(data = dfTraits %>% filter(!is.na(diet))) +
  # Barplot
  geom_bar(mapping = aes(x = diet, fill = diet), width = 0.7) +
  # Custom colours
  scale_fill_brewer(palette = "Paired") +
  labs(title = "Diet Types in Reptiles",
       x = "Diet",
       y = "Count") +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        legend.position = "none")

?scale_fill_brewer ## Lots of options, including colour blind friendly options

# Relationships between numerical variables
plot(dfTraits[contTraits]) ## scatter plots for each pair of traits

# Correlations
# Ranges from -1 to 1 and gives insight about the strength of pairwise relationships
?cor
# Note that you have the option to use different coefficients (Pearson, Kendall, Spearman)
cor(dfTraits[contTraits], use = "pairwise.complete.obs")

# Test for significant association between two traits
cor.test(dfTraits$hatchling_neonate_svl, dfTraits$age_first_breeding)

# The output tells us there is a statistically significant correlation (p-value < 0.05) between these two variables.

# Quick ggplot to see relationship
ggplot(data = dfTraits) +
  geom_point(mapping = aes(x = age_first_breeding, y = hatchling_neonate_svl))

# Are these data normally distributed? We can  this using QQ plots.
hist(dfTraits$hatchling_neonate_svl)
# The x-axis is the quantiles from the theoretical distribution we are comparing to (i.e., normal distribution) and y-axis is the quantiles from our data
qqnorm(dfTraits$hatchling_neonate_svl)
qqline(dfTraits$hatchling_neonate_svl) ## skewed distribution

# Let' see if we can use a log-transformation to make our data resemble a normal distribution
hist(dfTraits$hatchling_neonate_svl)
hist(log(dfTraits$hatchling_neonate_svl))
qqnorm(log(dfTraits$hatchling_neonate_svl))
qqline(log(dfTraits$hatchling_neonate_svl)) ## it helps!

# If you wanted to keep the original data in your dataset, you could use the mutate function to create a new column with the log-transformed data
dfTraits <- dfTraits %>%
  mutate(log_hatchling_neonate_svl = log(hatchling_neonate_svl),
         log_age_first_breeding = log(age_first_breeding))
names(dfTraits)

# Plot the transformed data.
# Note I placed the x and y variables in the ggplot argument here so I don't have to do it twice for both geom_point and geom_smooth
ggplot(data = dfTraits, aes(x = log_hatchling_neonate_svl, y = log_age_first_breeding)) +
  geom_point(# Adding some colour and transparency (alpha) to the points
             color = "skyblue", size = 2, alpha = 0.7) +
  # Add linear regression line
  geom_smooth(method = "lm", color = "darkblue", linewidth = 0.5) +  
  theme_minimal(base_size = 12)

# Boxplots to show distributions of SVL by IUCN redlist assessment
ggplot(data = dfTraits, mapping = aes(x = iucn_redlist_assessment, y = log(maximum_svl))) +
  # Boxplot
  geom_boxplot(aes(fill = iucn_redlist_assessment)) +
  scale_fill_brewer(palette = "Blues") +
  labs(title = "SVL Distribution by IUCN Redlist Assessment in Reptiles",
       x = "IUCN Redlist Assessment",
       y = "Log Maximum SVL") +
  # Flipping the coordinates for better visualization
  coord_flip() +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none") ## remove the legend

# Associations between categorical variables.
ggplot(data = dfTraits %>% filter(!is.na(diet))) +
  # Using a variant of geom_point
  geom_count(mapping = aes(x = diet, y = iucn_redlist_assessment)) +
  labs(x = "Diet",
       y = "IUCN Redlist Assessment") +
  # Increasing the size of the points
  scale_size_continuous(range = c(2, 10), name = "Count") +
  theme_minimal(base_size = 12) 
```

Based on this, you may ask yourself what variables/relationships you want to explore further and which you could possibly remove from your dataset.

### Outlier detection. ----

Remember those really long species? Let's look at them a bit more closely! But this time, we will group by family and highlight outliers using the outlier arguments available in geom_boxplot.

```{r, message=FALSE, warning=FALSE}
# Boxplots to show distributions of maximum SVL by family
ggplot(data = dfTraits %>% filter(!is.na(maximum_svl))) +
  geom_boxplot(mapping = aes(x = family, y = maximum_svl),
               outlier.color = "red") +
  coord_flip()
```

##### You can see the outliers are a part of the Varanidae family. 

```{r, message=FALSE, warning=FALSE}
# Using the interquartile method to identify outliers.
# Determine the 1st quartile using the quantile function.
quantile(dfTraits[, "maximum_svl"], na.rm = T)
lowerQuantile <- quantile(dfTraits[, "maximum_svl"], na.rm = T)[2]

# Determine the 3rd quartile using the quantile function.
upperQuantile <- quantile(dfTraits[, "maximum_svl"], na.rm = T)[4]
upperQuantile

# Calculate the IQR by subtracting the 1st quartile from the 3rd quartile.
iqr <- upperQuantile - lowerQuantile

# Calculate our upper threshold ((3 x the IQR) + upperQuantile).
upperThreshold <- (iqr * 3) + upperQuantile

# Identify outliers based on whether they exceed the upper threshold.
outliers <- which(dfTraits[, "maximum_svl"] > upperThreshold)

# Subset the outliers with taxonomic information.
dfOutliers <- dfTraits[outliers, c("family", "genus", "species", "maximum_svl")]

head(dfOutliers)
```

Detect points that may be need to be ed out further. Should you remove them or not? E.g., look further into whether they might be human error or a legitimate biological observation! And of course, the choice to remove outliers will also depend on the model you end up using.

### Dimensionality reduction. ----

If you have a high dimensional dataset, you may think about using PCA to make your data more manageable.

Principal component analysis example.

```{r, message=FALSE, warning=FALSE}
# Create a data subset for PCA analysis.
colnames(dfTraits)
# We need to remove NAs before doing this (but don't worry, we will deal with missing values this afternoon)!
dfPCA <- as.data.frame(na.omit(dfTraits[contTraits]))

# Perform PCA analysis using prcomp. 
pcaRes <- prcomp(dfPCA, center = T, scale = T)

# Standard deviations for each principal component.
pcaRes$sdev

# The columns here are eigenvectors that correspond to each principal component, tells you how much each variable contributes to the component
pcaRes$rotation

# The transformed variables in the PCA space.
pcaRes$x

# How much variance is explained by each of the components?
summary(pcaRes)

# This plot shows the percentage of variance explained by each of the components.
fviz_eig(pcaRes, addlabels = T)

# This plot visualizes both the principal components and the original variables.
# Longer arrows indicate greater contribution
fviz_pca_var(pcaRes,
             # colour by contribution
             col.var = "contrib",
             gradient.cols = "Paired",
             repel = T, 
             xlab = "Principal Component 1",
             ylab = "Principal Component 2") + ## takes ggplot arguments
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))

# Next step would be to extract the components to use in an analysis down the road (covered in later modules)

# From here, keep the variables you are interested in exploring for your analysis.
# But beware! Imputation prep is next, and it's not so forgiving!
dfTraits <- dfTraits %>%
  select(family, genus, species, latitude, insular_endemic, maximum_svl,
         hatchling_neonate_svl, activity_time, diet, foraging_mode,
         reproductive_mode, largest_clutch, age_first_breeding, 
         iucn_redlist_assessment)

# Write dataset to file for next module.
write.csv(dfTraits, "dfTraits.csv", row.names = F)
```

